---
title: "Step Analysis, based on the Data of Portable Monitors"
output: html_document
---

## Loading and preprocessing the data

1. Load the data and libraries
```{r}
library(dplyr)
library(ggplot2)
raw = read.csv("activity.csv")
head(raw)
```

## What is mean total number of steps taken per day?
For this part of the assignment, we ignored the missing values in the dataset.

1. Calculate the total number of steps taken per day
Firstly, the dataset raw is grouped according to the date, stored in by_group dataset.
Then we calculate the sum of each group, i.e. the total steps taken per day, stored in ave_step.
```{r}
by_date = group_by(raw, date)
ave_step = summarise(by_date, dailySum = sum(steps, na.rm = TRUE))
head(ave_step)
```

2. Make a histogram of the total number of steps taken each day
```{r}
qplot(dailySum, data = ave_step, xlab = "Number of Daily Steps", 
      ylab = "Numer of Days", geom = "histogram",fill = "red")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
stepData = ave_step$dailySum
MEAN = mean(stepData, na.rm = TRUE)
MEDIAN = median(stepData, na.rm = TRUE)
rbind(c("Mean of Daily Steps",MEAN), c("Median of Daily Steps", MEDIAN))
```

## What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
Firstly, group the raw data according to the 5-minute interval
Then, calculate the mean of the interval across all days
Lastly, use ggplot to draw the line diagram.
```{r}
by_interval = group_by(raw[,c(1,3)], interval)
ave_stepByInterval = summarise(by_interval, intervalMean = mean(steps, na.rm = TRUE))

ggplot(data = ave_stepByInterval, aes(x=interval, y=intervalMean)) + geom_line() + xlab("5-minute interval") + ylab("average number of steps taken")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_step_interval = ave_stepByInterval[which.max(ave_stepByInterval$intervalMean),]
max_step_interval
```

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as \color{red}{\verb|NA|}NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs)

First of all, by using is.na, we know that there are no missing values in date columen and interval column of dataset raw
```{r}
sum(is.na(raw$date)) + sum(is.na(raw$interval))
```
Missing values are lying in column steps. That is also the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs)
```{r}
sum(is.na(raw$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset. 
NA denotes that no value is recorded in the 5-minute interval. In order to replace NA value and not bias the final result, we replace the NA value with the average number of steps across all intervals on that day.

First, we calculate the mean number of steps of a interval every day. In case that there are several days when no data are recorded whole day, we put 0 
```{r}
by_Date = summarise(group_by(raw[,1:2], date), Intervalmean=mean(steps,na.rm = TRUE))
by_Date[is.nan(by_Date$Intervalmean),2]=0
by_Date
```

Then, clone raw dataset to NewRaw, which is going to store the cleaned dataset (without NA)
After that, check the cells in steps colume whether it is NA. If TRUE, extract the date of that cell, and replace the NA value by the average interval steps on that day.

See that the NA values of the first 6 rows are replaced with 0
```{r}
NewRaw = raw
for (i in 1:length(raw$steps))
{
        if (is.na(raw[i,1])) 
        {
                day = raw[i,2]
                match = by_Date[day,2]
                NewRaw[i,1] = match
                
        }
}
head(NewRaw)
```

3. Make a histogram of the total number of steps taken each day 

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
by_date2 = group_by(NewRaw, date)
ave_step2 = summarise(by_date2, dailySum = sum(steps))
qplot(dailySum, data=ave_step2, xlab = "Number of Daily Steps", ylab = "Numer of Days", binwidth=1000)
```

Calculate the mean and median total number of steps taken per day. 
```{r}
stepData2 = ave_step2$dailySum
MEAN2 = mean(stepData2)
MEDIAN2 = median(stepData2)   
rbind(c("Mean of Daily Steps",MEAN2), c("Median of Daily Steps", MEDIAN2))
```


## Differences in activity patterns between weekdays and weekends
Firstly, we add a colume WeekDay to the raw dataset, stored it in rawNwkd.
Then subset weekday data from rawNwkd, group it according to the interval. Then calculate the mean corresponding steps across all days. Add an indicator column WD in the end
Then do the same to the weekend.
Next, combine two dataset and plot the picture
```{r}
rawNwkd=mutate(NewRaw, WeekDay= weekdays(as.Date(date)))
WEEKDAY = subset(rawNwkd, WeekDay %in% c("Monday","Tuesday","Wednesday","Thursday","Friday"))
WEEKDAY2 = group_by(WEEKDAY[,c(1,3)], interval)
WEEKDAY3 = summarise(WEEKDAY2, ave= mean(steps))
WEEKDAY3 = mutate(WEEKDAY3, WD = 1)

WEEKEND = subset(rawNwkd, WeekDay %in% c("Saturday","Sunday"))
WEEKEND2 = group_by(WEEKEND[,c(1,3)], interval)
WEEKEND3 = summarise(WEEKEND2, ave= mean(steps))
WEEKEND3 = mutate(WEEKEND3, WD = 0)

combined = rbind(WEEKDAY3,WEEKEND3)
ggplot(data = combined, aes(x=interval,y=ave)) + geom_line(colour="#FF9999") + 
        facet_grid(rows = vars(WD))

```


